Jane, John … Leslie? A Historical Method for Algorithmic Gender Prediction

### 2015-08-24 article.Rmd

Researchers who use this method need to be fully aware of its limitations, in particular its dependency on a state-defined gender binary. Inferring gender from personal names is a blunt tool to study a complex subject. Gender theorists and feminist scholars have spent decades unpacking the full range of meanings behind gender as an analytical concept, including the fluid relationship between biological sex and gender identity [Butler 1990, Butler 1993]. On its own, our method has little to say about crucial topics such as the intersectionality of gender with race, class, and power, the lives of transgender people, or what the categories "male" or "female" mean in any given historical and social context [Crenshaw 1991, McCall 2005]. The tool cannot reflect how an individual person self-identifies and how gender shapes their lived experience, since it relies on large datasets created by state agencies such as the U.S. Census Bureau. Much like the state’s definition of racial categories, governments impose a gender binary in their data collection that says more about existing power structures than the ways in which people self-identify. [Brubaker and Cooper 2000, 14-17, Pascoe 2009, Canaday 2011, Landrum 2014]. When using the predictions provided by this method, users must judge for themselves how they reflect the social and cultural practices of the time and place under study.

@@

Nevertheless, the methodology that we have described here does take advantage of the historical datasets available to scholars. The predictions it generates help to uncover patterns about gender where they were previously unknown, providing further grist to the mill of more traditional methods. Furthermore, the gender package is entirely open-source, and we have provided detailed documentation for its code and underlying data so that those who wish to build upon it can modify the package accordingly. For instance, future researchers might compile datasets of names that move beyond the gender binary imposed by government agencies. They could then incorporate these datasets into the gender package, replacing the "Male" and "Female" designations with a much wider spectrum of potential gender identities. By building as much flexibility and transparency into the method as possible, we hope that others can continue to make just these kinds of improvements.

As it currently stands, the gender package meets a methodological need. As of August 2015, the gender package has been downloaded more than 21,000 times and incorporated as a pedagogical tool within digital humanities courses [Jockers 2014b].^[The download counts come from the [download logs from RStudio's CRAN mirror](http://cran-logs.rstudio.com/) as tabulated by Gabor Csardi's [cranlogs package](http://cranlogs.r-pkg.org/). [Csardi 2015] RStudio is the only CRAN mirror to release its logs.)] Researchers have used it to uncover instances gender inequity in a range of different areas, from biases in students' teaching evaluations to the disparities between attendees and authors at the annual Digital Humanities Conference [Schmidt 2015, Weingart 2015]. These implementations illustrate the value of the method. Its major intervention is to algorithmically infer gender in textual sources in a way that takes into account change over time. Rather than trying to shoehorn existing computer science methods to fit humanities research, the gender package reverses this traditional disciplinary relationship. It uses a humanistic starting point---the historical relationship between gender and naming practices---to contribute a new method to computer science.

>>>>

Researchers who use this method need to be fully aware of its limitations, in particular its dependency on a state-defined gender binary. Inferring gender from personal names is a blunt tool to study a complex subject. Gender theorists and feminist scholars have spent decades unpacking the full range of meanings behind gender as an analytical concept, including the fluid relationship between biological sex and gender identity [Butler 1990, Butler 1993]. On its own, our method has little to say about crucial topics such as the intersectionality of gender with race, class, and power, the lives of transgender people, or what the categories "male" or "female" mean in any given historical and social context [Crenshaw 1991, McCall 2005]. The leap from datasets that contain records of sex at birth to gender has rightly been critiqued by scholars [Posner 2015, Westbrook and Saperstein 2015]. The tool cannot reflect how an individual person self-identifies and how gender shapes their lived experience, since it relies on large datasets created by state agencies such as the U.S. Census Bureau. Much like the state’s definition of racial categories, governments impose a gender binary in their data collection that says more about existing power structures than the ways in which people self-identify. [Brubaker and Cooper 2000, 14-17, Pascoe 2009, Canaday 2011, Landrum 2014]. When using the predictions provided by this method, users must judge for themselves how they reflect the social and cultural practices of the time and place under study.

@@

Nevertheless, the methodology that we have described here does take advantage of the datasets available to historians. While we agree with the category of critique that we have cited, we do not think that they should deter scholars from making use of our method. While such critiques point out genuine problems, they also seldom contain concrete proposals that can actually be implemented in historical research.^[Discussing exactly this problem of gender/sex binaries in large corpora, Penelope Eckert argues that "binary sex-class gets us pretty far" and that "at the moment" it is not "whether there's any way we can code large corpora for something subtler about gender than sex-class" [Eckert 2012].]  The predictions it generates help to uncover patterns about gender where they were previously unknown, providing further grist to the mill of more traditional methods. Furthermore, the gender package is entirely open-source, and we have provided detailed documentation for its code and underlying data so that those who wish to build upon it can modify the package accordingly. For instance, future researchers might compile datasets of names that move beyond the gender binary imposed by government agencies. They could then incorporate these datasets into the gender package, replacing the "male" and "female" designations with a much wider spectrum of potential gender identities.

As it currently stands, the gender package meets a methodological need. As of August 2015, the gender package has been downloaded more than 22,000 times and incorporated as a pedagogical tool within digital humanities courses [Jockers 2014b].^[The download counts come from the [download logs from RStudio's CRAN mirror](http://cran-logs.rstudio.com/) as tabulated by Gabor Csardi's [cranlogs package](http://cranlogs.r-pkg.org/). [Csardi 2015] RStudio is the only CRAN mirror to release its logs.)] Researchers have used it to uncover instances gender inequity in a range of different areas, from biases in students' teaching evaluations to the disparities between attendees and authors at the annual Digital Humanities Conference [Schmidt 2015, Weingart 2015]. These implementations illustrate the value of the method. Its major intervention is to algorithmically infer gender in textual sources in a way that takes into account change over time. Rather than trying to shoehorn existing computer science methods to fit humanities research, the gender package reverses this traditional disciplinary relationship. It uses a humanistic starting point---the historical relationship between gender and naming practices---to contribute a new method to computer science.
